---
title: "Analysis"
author: "Saad Usmani, Beau Britain, Shane Caldwell, Mackenzie Gray"
date: "November 4, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

## Introduction

\bigskip
\bigskip
\bigskip

This document analyzes the College Scorecard Dataset provided by the Federal Government from the years 1996 - 2016. The dataset includes over 1700+ variables with names of over 7000+ colleges in the United States. We will explore this dataset and look at interesting relationships between variables and colleges. We also provide an R Shiny App that lets a user put in personal application information and desired college preferences to look at colleges that would be a "best fit" based on provided information, as well as provide potential costs related to loan repayment and average income for a major after graduation.

The dataset is large, spanning over 15 years with 1700 variables. To make our analysis simpler, we chose around 81 variables to look at that relate to a college's descriptive statistics - like admission rates, average SAT/ACT score, average percentage of degrees awarded to a specific major, number of undergraduate students, tuition rates, private and public colleges, etc. We spend the rest of the document detailing useful relations and variables. 

## Cleaning Our Dataset

An average student would like to look at the most recent statistics for a college - so with over 15+ years of data, we only looked at the last 5 years of relevant college statistics. To do this, we merged the years 2010-2015 into one large college data csv file, then took the averages of every column for each College/University. We do this with the following code:

```{r}
df1<-read.csv(file='C:/Users/susmani/Documents/Year 5/Courses/Data Munging/GroupProject1/CollegeScorecard_Raw_Data_2017/CollegeScorecard_Raw_Data/MERGED_2010-15_selected_columns.csv', sep = ',')
df<-df1%>%
  group_by(UNITID, INSTNM, STABBR) %>%   
  summarise_all(funs(mean(.,na.rm=TRUE)))
write.csv(df, 'MERGED_2010-15_AVG.csv', row.names = FALSE)
```

The write.csv line mearly writes a csv file that can easily be looked at. 

### Assessing Quality of Data

Looking at our new table that now consists of only 99 variables, we can assess the quality of our data. 

